---
title: "PS3_Applied"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

output: 
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
urlcolor: blue
---
```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy = "styler")
knitr::opts_chunk$set(fig.height = 3, fig.width = 5)
```


```{r message=FALSE, warning=FALSE, include = FALSE}
library(stringi)
library(splitstackshape)
library(tidyverse)
library(testthat)
library(pryr)
library(lubridate)
library(ggplot2)
library(tidycensus)
library(leaflet)
```

This submission is our work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **RM** **VS**

Ryan late coins used this pset: 0. Late coins left: 0. 
Vera late coins used this pset: 0. Late coins left: 1. 

#Load and glimpse

## Load and glimpse
```{r}
gun_violence <- read_csv("gun-violence-data_01-2013_03-2018.csv")

glimpse(gun_violence)
```
1.1 There are 239,677 rows and 29 columns.
1.2 The variables are: 
incident_id
date,state
city_or_county
address,n_killed

n_injured
incident_url
source_url
incident_url_fields_missing
congressional_district
gun_stolen
gun_type
incident_characteristics - *this is very interesting as it gives a good overview over the incident.  It may be helpful to seperate this variable into more variables*
latitude,location_description
longitude,n_guns_involved,notes
participant_age
participant_age_group
participant_gender
participant_name
participant_relationship
participant_status
participant_type
sources
state_house_district
state_senate_district	

1.3. 
We may want to format the following variables as factors: gun_stolen, gun_type,incident_characteristics, participant_age and Pariticpant_age_group, participant_status and participant_type




## Explore
```{r}
head(gun_violence)
```



## Format

We might have to format the participant variables, gun stolen, gun type, incident chracteristics into factors


# Time

## Date
```{r}
gun_violence$date %>% 
  str()
```
It is aldready  in the date format.

## Increasing by year?

```{r}
gun_violence %>% 
mutate(year = year(date)) %>% 
  group_by(year) %>% 
  summarize(count = n()) %>% 
ggplot(aes(year, count)) +
  geom_bar(stat = "identity")
```

## month



```{r}
gun_violence %>% 
mutate(month = month(date, label = TRUE)) %>% 
mutate(year=year(date)) %>% 
  group_by(month,year) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(month, count)) +
    geom_bar(stat = "identity")+
    facet_wrap(~year)+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      theme(panel.spacing = unit(1, "lines"))

```
Generally speaking I think the summer months have more shoting incidents.  It is hard to distinguish between June, July and August


## Day

```{r}
gun_violence %>% 
  mutate(wday = wday(date, label = TRUE)) %>% 
  mutate(year=year(date)) %>%
  group_by(wday, year) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(wday, count)) +
    geom_bar(stat = "identity")+
    facet_wrap(~year)+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      theme(panel.spacing = unit(1, "lines"))+
  labs(x='Weekday', y='Number of incidents', title='Incidents by Weekday')
```

## Summary

More violence om weekends, more violence in the summer, and more violence over the years.


# Characteristics

## Avg number guns
```{r}
gun_violence %>% 
  filter(n_guns_involved != is.na(n_guns_involved)) 
  summarize(mean = mean(n_guns_involved))
```
100,000 rows are missing this data, and of the ones that recorded, there 1.37 guns per incident.

## Type of guns
Incredibly complicated but got there in the end- the graph needs to be be labelled etc

```{r}


    
  
gun_violence$gun_type <- gsub("\\|\\|", "|", gun_violence$gun_type)

gun_type_columnized <- splitstackshape::cSplit(gun_violence %>% select( gun_type), 'gun_type', sep =  '|', direction="long")

gun_type_columnized



gun_types_cleaned<- gun_type_columnized %>% 
  drop_na() %>% 
  filter(str_detect(gun_type,"Unknown", negate = TRUE)) %>%  
  mutate(gun=stri_sub(gun_type,4))

gun_types_cleaned

gun_types_cleaned %>%
  group_by(gun) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count) ) %>% 
  filter(count>=196) %>% 
  ggplot(aes(x=gun, count))+
  geom_histogram(stat="identity")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



    
```


##incident characteristics

got help from https://www.kaggle.com/erikbruin/gun-violence-in-the-us-eda-and-rshiny-app

```{r}


gun_violence$incident_characteristics <- gsub("\\|\\|", "|", gun_violence$incident_characteristics)

IncCharac <- splitstackshape::cSplit(gun_violence %>% select(incident_id, state, city_or_county, incident_characteristics), 'incident_characteristics', sep =  '|', direction="long")



numCat <- round(nrow(IncCharac)/nrow(gun_violence),1)
cat('On average, there are', numCat, 'incident categories specified per incident')

IncCharac %>% 
  head(10)

IncCharac %>% 
  group_by(incident_characteristics) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count) ) %>% 
  filter(count>6300) %>% 
  ggplot(aes(x=incident_characteristics, count))+
  geom_histogram(stat="identity")+
  theme(axis.text.x = element_text(angle = 55, vjust = 0.5, hjust=1))


```


# suspects characterisitics


By way of example let us look at the first row.
In the age group there are 5 individuals, in the gender group there are four etc.  It can be hard to reconcile the individuals

## Avg number suspects and victims
```{r}
gun_violence %>% 
  select(participant_age,participant_age_group,participant_gender,participant_name, participant_status,participant_type) %>% 
  head(1)
```

```{r,fig.width=10}


gun_filter= gun_violence %>% 
  mutate(year=year(date)) %>% 
  filter(year=='2017')

type=cSplit(gun_filter,c("participant_status"),sep="||",direction="long",drop=TRUE)


type

type$participant_status=gsub(".*:","",type$participant_status)




head(type$participant_status)

total=length(unique(type$incident_id))
temp = type %>% group_by(participant_status) %>% summarise(count=n()) %>% mutate(perc=round((count/total)*100,2)) %>% arrange(desc(count))


ggplot(temp,aes(factor(participant_status,participant_status),count,fill=participant_status))+geom_bar(stat="identity")+geom_text(aes(label=paste0(perc,"%",sep=" ")),hjust=0,vjust=0.5,size=4,color='black',fontface='bold')+labs(x="Participant Status",y="Count",title='Year:2017 - Participant Status')+coord_flip()
```





4.2 average of suspects and victims per incident
We choose to use the participant type and to chop it up to find the numer of participants

```{r}
gun_violence$participant_status <- gsub("\\|\\|", "|", gun_violence$participant_status)

gun_violence$participant_status

participant_status_columned <- splitstackshape::cSplit(gun_violence %>% select(incident_id,participant_status), 'participant_status', sep =  '|', direction="long")


participant_status_columned %>% 
  group_by(incident_id) %>% 
  summarise(count=n()) %>% 
  summarise(mean=mean(count))



```
the mean number of victims and suspects is 1.7 which must mean that sometimes there is just a victim and no suspect



4.3 Group by id, participant_gender and participant_age 



```{r}
Gender_participation <- gun_violence %>% 
  select(incident_id, participant_type, participant_age,participant_gender)


type=cSplit(Gender_participation,c("participant_type"),sep="||",direction="long",drop=TRUE)
type=cSplit(type,c("participant_age"),sep="||",direction="long",drop=TRUE)


type

type<-type %>% 
  head(10000) %>% 
  filter(str_detect(participant_type, "Subject-Suspect", negate=FALSE)) #picking only suspect
type

type<-type %>% 
  filter(substr(participant_type, 1, 3)==substr(participant_age,1,3))  # chose rows were the number of suspect and age match

type<-type %>% 
  mutate(suspect_age=as.numeric(substr(participant_age,4,5))) # put suspect age into a separate column
 
type




```
4.4 show age distribution  - this graph needs labels etc.

```{r}
type %>% 
  filter(suspect_age<100) %>% 
  ggplot(aes(suspect_age))+
  geom_bar()
```
4.5 Gender distribution


```{r}

Gender_participation <- gun_violence %>% 
  select(incident_id, participant_type, participant_age,participant_gender)


type=cSplit(Gender_participation,c("participant_type"),sep="||",direction="long",drop=TRUE)
type=cSplit(type,c("participant_gender"),sep="||",direction="long",drop=TRUE)


type

type<-type %>% 
  head(10000) %>% 
  filter(str_detect(participant_type, "Subject-Suspect", negate=FALSE)) #picking only suspect
type

type<-type %>% 
  filter(substr(participant_type, 1, 3)==substr(participant_gender,1,3))  # chose rows were the number of suspect and age match

type<-type %>% 
  mutate(suspect_gender=(substr(participant_gender,4,7))) # put suspect gender into a separate column

type


type %>% 
  drop_na(suspect_gender) %>% 
  group_by(suspect_gender) %>% 
    summarize(count = n())
```
Out of the 1000 incidents we picked the there are 519 suspects.  Male are 499/519 = 96%. 

4.6. participant statuses

```{r}
c_participants<- gun_violence%>%   select(participant_status) %>% 
  cSplit(c("participant_status"),sep="|",direction="long",drop=TRUE) %>% 
  mutate(clean_status=substr(participant_status,4,50)) 

c_participants

length(unique(c_participants$clean_status))


```
The number of unique statuses is 28

4.7
```{r}
type <- gun_violence %>% 
  select(incident_id, participant_type, participant_status,participant_gender)


type=cSplit(type,c("participant_type"),sep="||",direction="long",drop=TRUE)
type=cSplit(type,c("participant_status"),sep="||",direction="long",drop=TRUE)


type

type<-type %>% 
  head(10000) %>% 
  filter(str_detect(participant_type, "Subject-Suspect", negate=FALSE)) #picking only suspect
type

type<-type %>% 
  filter(substr(participant_type, 1, 3)==substr(participant_status,1,3)) %>%   # chose rows were the number of suspect and status match
  filter(str_detect(participant_status,"Arrested"))
type

type<-type %>% 
  mutate(suspect_age=as.numeric(substr(participant_age,4,5))) # put suspect age into a separate column
 

```
Out of the 10 000 cases we looked at, a suspect had been arrested in 165 instances



#6. geographic variation


```{r}
gun_violence_dates <- gun_violence %>% 
mutate(year = year(date),
       month = month(date, label = TRUE),
       wday = wday(date, label = TRUE))


census_api_key("7b144a5d849b2f47475f9c1e44514c692f0246d0")

acs_data <- get_acs(
  geography = "state",
  variables = c("B01003_001"),
year = 2017
) %>% 
  arrange(NAME)

acs_data <- acs_data %>% 
  rename("state" = NAME,
         "population_est" = estimate)
	
gun_merged <- gun_violence_dates %>% 
  left_join(acs_data, by = "state")


gun_merged %>% 
  filter(year == 2017) %>% 
  group_by(state) %>% 
  summarize(count = n(),
            per_100 = (count/population_est)) %>% 
  ggplot(aes(reorder(state, per_100), per_100)) +
  geom_bar(stat = "identity", fill = "red") +
  labs(x = "State", y = "Number of Incidents", title = "States") +
  coord_flip()







```

##results in a map

THIS IS NOT FINISHED AND IS unlikely to be


```{r}


top_map<- gun_merged %>% 
  filter(year == 2017) %>% 
  group_by(state,longitude,latitude) %>% 
  drop_na(longitude) %>% 
  summarize(count = n(),
    per_100 = (count/population_est)) %>% 
  arrange(desc(per_100)) 

top_map<-top_map[!duplicated(top_map$state), ]

 
top_map



labels <- paste0("<strong>State: </strong>", top_map$state, 
                 "<br><strong>Victims </strong>", top_map$per_100) %>%    lapply(htmltools::HTML)

leaflet(top_map) %>%
        setView(lng=-96, lat=37.8, zoom=4) %>%
        addTiles() %>%
        addProviderTiles("CartoDB.Positron") %>%
        addCircleMarkers(~longitude, ~latitude, color = "red", radius=~sqrt(top_map), label = labels)

```





## Incidents with highest numbers of victims
We used bits from https://www.kaggle.com/erikbruin/gun-violence-in-the-us-eda-and-rshiny-app



```{r}



gun_violence$victims <- gun_violence$n_killed + gun_violence$n_injured # new variable created

Top15 <- gun_violence %>% 
  select(incident_id, date, n_killed, n_injured, victims, location_description, city_or_county, state, latitude, longitude) %>%
  rename(Incident_Id=incident_id, Date=date, Killed=n_killed, Injured=n_injured, Victims=victims, Location=location_description, City=city_or_county) %>%
  arrange(desc(Victims)) %>% 
  top_n(n=15, wt=Victims)

Top15

TopMap <- Top15 %>% select(latitude, longitude, Victims, state, Location)

labels <- paste0("<strong>State: </strong>", TopMap$state, 
                 "<br><strong>Location: </strong>", TopMap$Location,
                 "<br><strong>Victims </strong>", TopMap$Victims) %>%    lapply(htmltools::HTML)

leaflet(TopMap) %>%
        setView(lng=-96, lat=37.8, zoom=4) %>%
        addTiles() %>%
        addProviderTiles("CartoDB.Positron") %>%
        addCircleMarkers(~longitude, ~latitude, color = "red", radius=~sqrt(Victims), label = labels)



```

```{r}
kable(Top10 %>% select(-longitude, -latitude))
```

### An interactive map of the incidents with highest numbers of victims

```{r, out.width="100%"}
TopMap <- Top10 %>% select(latitude, longitude, Victims, City, Location)

labels <- paste0("<strong>City: </strong>", TopMap$City, 
                 "<br><strong>Location: </strong>", TopMap$Location,
                 "<br><strong>Victims </strong>", TopMap$Victims) %>% lapply(htmltools::HTML)

leaflet(TopMap) %>%
        setView(lng=-96, lat=37.8, zoom=4) %>%
        addTiles() %>%
        addProviderTiles("CartoDB.Positron") %>%
        addCircleMarkers(~longitude, ~latitude, color = "red", radius=~sqrt(Victims), label = labels)

# mass shootings
