---
title: "PS3_Applied"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

output: 
  pdf_document:
    number_sections: yes
  html_document:
    df_print: paged
urlcolor: blue
---
```{r, include = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
knitr::opts_chunk$set(tidy = "styler")
knitr::opts_chunk$set(fig.height = 3, fig.width = 5)
```


```{r message=FALSE, warning=FALSE, include = FALSE}
library(stringi)
library(splitstackshape)
library(tidyverse)
library(testthat)
library(pryr)
library(lubridate)
library(ggplot2)
library(tidycensus)
```

This submission is our work alone and complies with the 30535 integrity policy.

Add your initials to indicate your agreement: **RM** **VS**

Ryan late coins used this pset: 0. Late coins left: 0. 
Vera late coins used this pset: 0. Late coins left: 1. 

#Load and glimpse

## Load and glimpse
```{r}
gun_violence <- read_csv("gun-violence-data_01-2013_03-2018.csv")

glimpse(gun_violence)
```
1.1 There are 239,677 rows and 29 columns.
1.2 The variables are: 
incident_id
date,state
city_or_county
address,n_killed

n_injured
incident_url
source_url
incident_url_fields_missing
congressional_district
gun_stolen
gun_type
incident_characteristics - *this is very interesting as it gives a good overview over the incident.  It may be helpful to seperate this variable into more variables*
latitude,location_description
longitude,n_guns_involved,notes
participant_age
participant_age_group
participant_gender
participant_name
participant_relationship
participant_status
participant_type
sources
state_house_district
state_senate_district	

1.3. 
We may want to format the following variables as factors: gun_stolen, gun_type,incident_characteristics, participant_age and Pariticpant_age_group, participant_status and participant_type




## Explore
```{r}
head(gun_violence)
```



## Format

We might have to format the participant variables, gun stolen, gun type, incident chracteristics into factors


# Time

## Date
```{r}
gun_violence$date %>% 
  str()
```
It is aldready  in the date format.

## Increasing by year?

```{r}
gun_violence %>% 
mutate(year = year(date)) %>% 
  group_by(year) %>% 
  summarize(count = n()) %>% 
ggplot(aes(year, count)) +
  geom_bar(stat = "identity")
```

## month



```{r}
gun_violence %>% 
mutate(month = month(date, label = TRUE)) %>% 
mutate(year=year(date)) %>% 
  group_by(month,year) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(month, count)) +
    geom_bar(stat = "identity")+
    facet_wrap(~year)+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      theme(panel.spacing = unit(1, "lines"))

```
Generally speaking I think the summer months have more shoting incidents.  It is hard to distinguish between June, July and August


## Day

```{r}
gun_violence %>% 
  mutate(wday = wday(date, label = TRUE)) %>% 
  mutate(year=year(date)) %>%
  group_by(wday, year) %>% 
  summarize(count = n()) %>% 
  ggplot(aes(wday, count)) +
    geom_bar(stat = "identity")+
    facet_wrap(~year)+
      theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))+
      theme(panel.spacing = unit(1, "lines"))+
  labs(x='Weekday', y='Number of incidents', title='Incidents by Weekday')
```

## Summary

More violence om weekends, more violence in the summer, and more violence over the years.


# Characteristics

## Avg number guns
```{r}
gun_violence %>% 
  filter(n_guns_involved != is.na(n_guns_involved)) 
  summarize(mean = mean(n_guns_involved))
```
100,000 rows are missing this data, and of the ones that recorded, there 1.37 guns per incident.

## Type of guns
Incredibly complicated but got there in the end- the graph needs to be be labelled etc

```{r}


    
  
gun_violence$gun_type <- gsub("\\|\\|", "|", gun_violence$gun_type)

gun_type_columnized <- splitstackshape::cSplit(gun_violence %>% select( gun_type), 'gun_type', sep =  '|', direction="long")

gun_type_columnized



gun_types_cleaned<- gun_type_columnized %>% 
  drop_na() %>% 
  filter(str_detect(gun_type,"Unknown", negate = TRUE)) %>%  
  mutate(gun=stri_sub(gun_type,4))

gun_types_cleaned

gun_types_cleaned %>%
  group_by(gun) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count) ) %>% 
  filter(count>=196) %>% 
  ggplot(aes(x=gun, count))+
  geom_histogram(stat="identity")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))



    
```


##incident characteristics

got help from https://www.kaggle.com/erikbruin/gun-violence-in-the-us-eda-and-rshiny-app

```{r}


gun_violence$incident_characteristics <- gsub("\\|\\|", "|", gun_violence$incident_characteristics)

IncCharac <- splitstackshape::cSplit(gun_violence %>% select(incident_id, state, city_or_county, incident_characteristics), 'incident_characteristics', sep =  '|', direction="long")



numCat <- round(nrow(IncCharac)/nrow(gun_violence),1)
cat('On average, there are', numCat, 'incident categories specified per incident')

IncCharac %>% 
  head(10)

IncCharac %>% 
  group_by(incident_characteristics) %>% 
  summarise(count=n()) %>% 
  arrange(desc(count) ) %>% 
  filter(count>6300) %>% 
  ggplot(aes(x=incident_characteristics, count))+
  geom_histogram(stat="identity")+
  theme(axis.text.x = element_text(angle = 55, vjust = 0.5, hjust=1))


```


# suspects characterisitics


By way of example let us look at the first row.
In the age group there are 5 individuals, in the gender group there are four etc.  It can be hard to reconcile the individuals

## Avg number suspects and victims
```{r}
gun_violence %>% 
  select(participant_age,participant_age_group,participant_gender,participant_name, participant_status,participant_type) %>% 
  head(1)
```

```{r,fig.width=10}


gun_filter= gun_violence %>% 
  mutate(year=year(date)) %>% 
  filter(year=='2017')

type=cSplit(gun_filter,c("participant_status"),sep="||",direction="long",drop=TRUE)


type

type$participant_status=gsub(".*:","",type$participant_status)




head(type$participant_status)

total=length(unique(type$incident_id))
temp = type %>% group_by(participant_status) %>% summarise(count=n()) %>% mutate(perc=round((count/total)*100,2)) %>% arrange(desc(count))


ggplot(temp,aes(factor(participant_status,participant_status),count,fill=participant_status))+geom_bar(stat="identity")+geom_text(aes(label=paste0(perc,"%",sep=" ")),hjust=0,vjust=0.5,size=4,color='black',fontface='bold')+labs(x="Participant Status",y="Count",title='Year:2017 - Participant Status')+coord_flip()
```





4.2 average of suspects and victims per incident
We choose to use the participant type and to chop it up to find the numer of participants

```{r}
gun_violence$participant_status <- gsub("\\|\\|", "|", gun_violence$participant_status)

gun_violence$participant_status

participant_status_columned <- splitstackshape::cSplit(gun_violence %>% select(incident_id,participant_status), 'participant_status', sep =  '|', direction="long")


participant_status_columned %>% 
  group_by(incident_id) %>% 
  summarise(count=n()) %>% 
  summarise(mean=mean(count))



```
the mean number of victims and suspects is 1.7 which must mean that sometimes there is just a victim and no suspect


#geographic variation

# mass shootings
